---
title: "Final Project - Practical Machine Learning (Coursera)"
author: "Maanas Mutreja"
date: "11 September 2016"
output: html_document
---
##Introduction
This is a report produced during a homework assignment of Coursera's MOOC Practical Machine Learning from Johns Hopkins University.
The scripts have been solely produced, tested and executed on Windows 10, and RStudio Version 3.3.1


##Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here:http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


##Data Sources
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project comes from this original source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

Please Note that the data was downloaded and then used for analysis.


##Goal
The goal of the project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. We may use any of the other variables to predict with. We should create a report describing how you built your model, how we used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.
1. The submission should consist of a link to a Github repo with the R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5.
2. We should also apply the machine learning algorithm to the 20 test cases available in the test data above. Please submit the predictions in appropriate format to the programming assignment for automated grading.


##Initial Note
The initial loading of data to memory involves assuming some values (NA, #DIV/0! and blank values) as NA using "na.strings" argument while reading the data.


##Loading all relevant Libraries
```{r echo=TRUE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
```


##Getting and Loading the Data
The data is loaded from the local drive after downloading the data from URL.
```{r}
set.seed(12345)
setwd("F:/Personal/Coursera/Data Science/Prac Machine Lrnng/W4/Project")
training <- read.csv("training.csv", na.strings=c("NA","#DIV/0!",""))
testing <- read.csv("test.csv", na.strings=c("NA","#DIV/0!",""))
```


##Partition the Training set into two
```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining); dim(myTesting)
rm(inTrain)
```


##Cleaning the Data
Various Transformations are required to clean this data for analysis. Each transformation has been enumerated.

Transformation 1 - Remove NearZeroVariance variables
```{r}
nzv <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(myTesting,saveMetrics=TRUE)
myTesting <- myTesting[,nzv$nzv==FALSE]
rm(nzv)
```

Transformation 2 - Remove the First Column of myTraining dataset
```{r}
myTraining <- myTraining[c(-1)]
```

Transformation 3 - Clean variables with more than 60% NA
```{r}
trainingV3 <- myTraining
for(i in 1:length(myTraining)) {
    if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) == 1)  {
                trainingV3 <- trainingV3[ , -j]
            }   
        } 
    }
}

# Set back to the original variable name
myTraining <- trainingV3
rm(trainingV3)
```

Transformation 4 - Transform the myTesting and testing data sets
```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])  # remove the classe column
myTesting <- myTesting[clean1]         # allow only variables in myTesting that are also in myTraining
testing <- testing[clean2]             # allow only variables in testing that are also in myTraining
rm(clean1, clean2)
dim(myTesting)
```
```{r}
dim(testing)
```

Transformation 5 - Coerce the data into the same type
```{r}
for (i in 1:length(testing) ) {
    for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}

# To get the same class between testing and myTraining
testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```


##Using ML algorithms for prediction: Decision Tree
```{r}
set.seed(12345)
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)
```

```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
cmtree <- confusionMatrix(predictionsA1, myTesting$classe)
cmtree
```

```{r}
plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```


##Using ML algorithms for prediction: Random Forests
```{r}
set.seed(12345)
modFitB1 <- randomForest(classe ~ ., data=myTraining)
predictionB1 <- predict(modFitB1, myTesting, type = "class")
cmrf <- confusionMatrix(predictionB1, myTesting$classe)
cmrf
```

```{r}
plot(modFitB1)
```

```{r}
plot(cmrf$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```


##Using ML algorithms for prediction: Generalized Boosted Regression
```{r}
set.seed(12345)
fitControl <- trainControl(method = "repeatedcv", number = 5, repeats = 1)

gbmFit1 <- train(classe ~ ., data=myTraining, method = "gbm", trControl = fitControl, verbose = FALSE)

gbmFinMod1 <- gbmFit1$finalModel

gbmPredTest <- predict(gbmFit1, newdata=myTesting)
gbmAccuracyTest <- confusionMatrix(gbmPredTest, myTesting$classe)
gbmAccuracyTest
```

```{r}
plot(gbmFit1, ylim=c(0.9, 1))
```


##Prediction Results on test Data
Random Forests gave an Accuracy in the myTesting dataset of 99.89%, which was more accurate that what I got from the Decision Trees or GBM. The expected out-of-sample error is 100-99.89 = 0.11%.
```{r}
predictionB2 <- predict(modFitB1, testing, type = "class")
predictionB2
```

```{r}
# Write the results to a text file for submission
pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}

pml_write_files(predictionB2)
```








